---
title: "Lectures 3 - 4"
author: "Juan Carlos Villaseñor-Derbez"
date: "12/8/2020"
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(rethinking)
  library(brms)
  library(tidyverse)
  library(magrittr)
})

# Load some data
data("Howell1")
d <- Howell1

d2 <- d %>% 
  filter(age >= 18)
```

Define the model

$$
\begin{align}
h_i &\sim Normal(\mu, \sigma) \\
\mu &\sim Normal(178, 20)\\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

# Simulate from both priors to get a prior probability distribution of heights

```{r}
n <- 1e4                                                # We want 10,000 points

# Create a data.frame with dandom sample sof mu and sigma
sim <-
  tibble(sample_mu = rnorm(n = n, mean = 178, sd = 20), 
         sample_sigma = runif(n = n, min = 0, max = 50)) %>% 
  mutate(height = rnorm(n = n, mean = sample_mu, sd = sample_sigma))           #Draw random samples

# Plot it
ggplot(data = sim, mapping = aes(x = height)) +
  geom_density(fill = "gray") +
  labs(title = "height ~ dnrom(mu, sigma)") +
  theme_minimal()
```

# Use grid approximation of the posterior

```{r}
n <- 200                              # We'll use a 200 X 200 grid

# Create a function
grid_function <- function(mu, sd, data){
  sum(
    dnorm(x = data, mean = mu, sd = sd, log = TRUE)
  )
}

# Pull a vector to use
heights <- pull(d2, height)

# Create the grid, and call the function
d_grid <- 
  crossing(mu = seq(from = 140, to = 160, length.out = n),
           sigma = seq(from = 4, to = 9, length.out = n)) %>% 
  mutate(log_likelihood = map2_dbl(mu, sigma, grid_function, data = heights),
         prior_mu  = dnorm(x = mu, mean = 178, sd = 20, log = TRUE),
         prior_sigma = dunif(x = sigma, min = 0, max = 50, log = TRUE),
         prod = log_likelihood +  prior_mu + prior_sigma,
         prob = exp(prod - max(prod))) 
```

## Plot this grid

```{r}
ggplot(data = d_grid, mapping = aes(x = mu, y = sigma, z = prob, fill = prob)) +
  geom_raster(alpha = 0.8) +
  geom_contour(color = "black") +
  scale_fill_viridis_c() +
  theme_minimal()
```

# Sample from the posterior

```{r, fig.width = 4, fig.height = 4}

d_grid_samples <- 
  d_grid %>% 
  sample_n(size = 1e4, replace = T, weight = prob)

ggplot(data = d_grid_samples,
       mapping = aes(x = mu, y = sigma)) +
  geom_point(size = 1, alpha = 0.1) +
  theme_minimal()
```

Let's inspect the posteriors of each parameter

```{r}
d_grid_samples %>% 
  pivot_longer(mu:sigma) %>% 
  ggplot(aes(x = value)) +
  geom_density(fill = "gray") +
  facet_wrap(~name, scales = "free") +
  theme_minimal()
```

# Estimate the posterior with functions

First using `quap`

```{r}
flist <- alist(
  height ~ dnorm(mu, sigma),
  mu ~ dnorm(178, 20), 
  sigma ~ dunif(0, 50)
)

model <- quap(flist, data = d2)

precis(model)

```
And now using brms::brm

```{r}
model <- 
  brm(formula = height ~ 1, data = d2, family = gaussian,
      prior = c(prior(normal(178, 20), class = Intercept),
                prior(cauchy(0, 1), class = sigma)),
      iter = 2000,
      warmup = 1000,
      chains = 4,
      cores = 4)
```


Inspect the chains

```{r}
plot(model)
```

Lets use tighter priors

```{r}
model_2 <- 
  brm(formula = height ~ 1, data = d2, family = gaussian,
      prior = c(prior(normal(178, 1), class = Intercept),
                prior(cauchy(0, 1), class = sigma)),
      iter = 2000,
      warmup = 1000,
      chains = 4,
      cores = 4)
```

```{r}
plot(model_2)
```

## Sampling from model fits


```{r}
# Get the HMC chains out into a data.frame
post <- posterior_samples(model) %>% 
  select(b_Intercept, sigma)

post %>% 
  cov() %>% 
  diag()

post %>% 
  cor()

```

```{r}
posterior_summary(model)
```



# Now a linear model

$$
\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \\
\mu_i &= a + \beta(x_i - \bar{x})\\
\alpha &\sim Normal(178, 20)\\
\beta &\sim Normal(0, 10) \\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

## Simulate priors

```{r}
set.seed(2971)
n_lines <- 100

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b = rnorm(n = n_lines, mean = 0, sd = 10)) %>% 
  expand(nesting(n, a, b), weight = range(d2$weight)) %>% 
  mutate(height = a + b * (weight - mean(d2$weight))) %>% 
  ggplot(aes(x = weight, y = height, group = paste(a, b))) +
  geom_line(size = 0.1) +
  theme_bw() +
  labs(title = "b ~ dnorm(0, 10)")
```


Lets modify the prior, because the prior simulations tell us it is shit.

$$
\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \\
\mu_i &= a + \beta(x_i - \bar{x})\\
\alpha &\sim Normal(178, 20)\\
\beta &\sim Log-Normal(0, 1) \\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

## Simulate priors again

```{r}
set.seed(2971)
n_lines <- 100

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b = rlnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b), weight = range(d2$weight)) %>% 
  mutate(height = a + b * (weight - mean(d2$weight))) %>% 
  ggplot(aes(x = weight, y = height, group = paste(a, b))) +
  geom_line(size = 0.1) +
  theme_bw() +
  labs(title = "b ~ dnorm(0, 10)")
```


## Approximate the posterior using brms

```{r}

d2 %<>%
  mutate(weight_c = weight - mean(weight, na.rm = T))

model <- 
  brm(formula = height ~ 1 + weight_c,
      data = d2,
      family = gaussian,
      prior = c(prior(normal(178, 20), class = Intercept),
                prior(lognormal(0, 1), class = b),
                prior(uniform(0, 50), class = sigma)),
      iter = 28000, warmup = 27000, chains = 4, cores = 8,
      seed = 4,
      file = "my_model"
      )

plot(model)
```
```{r}
posterior_samples(model)[, 1:3] %>% 
  cov() %>% 
  round(digits = 3)
```

```{r}
pairs(model)
```
## Plot predictions

```{r}
ggplot(data = d2, mapping = aes(x = weight_c, y = height)) +
  geom_point() +
  geom_abline(intercept = fixef(model)[1],
              slope = fixef(model)[2])
```

### Add uncertainty around the mean

```{r}
weight_seq <- 
  tibble(weight = 25:70) %>% 
  mutate(weight_c = weight - mean(d2$weight))

mu_summary <- 
  fitted(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)


ggplot(data = d2, mapping = aes(x = weight_c)) +
  geom_point(aes(y = height)) +
  geom_smooth(data = mu_summary,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "red",
              size = 1,
              alpha = 0.5,
              color = "black") +
  theme_bw()
```

### Add uncertainty around the entire prediction

```{r}
pred_height <- 
  predict(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

ggplot(data = d2, mapping = aes(x = weight_c)) +
  geom_point(aes(y = height)) +
  geom_ribbon(data = pred_height,
              aes(ymin = Q2.5, ymax = Q97.5),
              alpha = 0.5) +
  geom_smooth(data = mu_summary,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "red",
              size = 1,
              alpha = 0.5,
              color = "black") +
  theme_bw()
```


# Quadratic model

$$
\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \\
\mu_i &= a + \beta_1weight\_s_i + \beta_2weight\_s_i^2\\
\alpha &\sim Normal(178, 20)\\
\beta_1 &\sim Log-Normal(0, 1) \\
\beta_2 &\sim Normal(0, 1) \\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

## Simulating from the priors

```{r}
set.seed(2971)
n_lines <- 100

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b_1 = rlnorm(n = n_lines, mean = 0, sd = 1),
       b_2 = rnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b_1, b_2), weight = seq(4, 63, by = 5)) %>% 
  mutate(height = a + b_1* (weight - mean(d$weight)) + (b_2 * (weight - mean(d$weight)) ^ 2)) %>% 
  ggplot(aes(x = weight, y = height)) +
  geom_point(data = d) +
  geom_line(size = 0.1, aes(group = n)) +
  theme_bw() +
  labs(title = "b ~ dnorm(0, 10)")
```

## Fit the model

```{r}
d %<>%
  mutate(weight_s = (weight - mean(weight, na.rm = T)) / sd(weight, na.rm = T),
         weight_s2 = weight_s ^ 2)

model <- brm(
  formula = height ~ 1 + weight_s + weight_s2,
  data = d,
  family = gaussian,
  prior = c(prior(normal(178, 100), class = Intercept),
            prior(lognormal(0, 1), class = b, coef = "weight_s"),
            prior(normal(0, 1), class = b, coef = "weight_s2"),
            prior(uniform(0, 50), class = sigma)),
  iter = 30000, warmup = 29000, chains = 4, cores = 4,
  seed = 4
)

```

```{r}
plot(model)
```
## And plot it

```{r}
weight_seq <- 
  tibble(weight_s = seq(from = -2.5, to = 2.5, length.out = 30)) %>% 
  mutate(weight_s2 = weight_s ^ 2)

mu_summary <- 
  fitted(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

pred_height <- 
  predict(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

ggplot(data = d, mapping = aes(x = weight_s)) +
  geom_point(aes(y = height)) +
  geom_ribbon(data = pred_height,
              aes(ymin = Q2.5, ymax = Q97.5),
              alpha = 0.5) +
  geom_smooth(data = mu_summary,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "red",
              size = 1,
              alpha = 0.5,
              color = "black") +
  theme_bw()
```

# Now a cubic model

$$
\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \\
\mu_i &= a + \beta_1weight\_s_i + \beta_2weight\_s_i^2 + \beta_3weight\_s_i^3\\
\alpha &\sim Normal(178, 20)\\
\beta_1 &\sim Log-Normal(0, 1) \\
\beta_2 &\sim Normal(0, 1) \\
\beta_3 &\sim Normal(0, 1) \\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

## Simulate the priors

```{r}
set.seed(2971)
n_lines <- 100

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b_1 = rlnorm(n = n_lines, mean = 0, sd = 1),
       b_2 = rnorm(n = n_lines, mean = 0, sd = 1),
       b_3 = rnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b_1, b_2, b_3), weight = seq(4, 63, length.out = 20)) %>% 
  mutate(height = a + b_1* (weight - mean(d$weight)) + (b_2 * (weight - mean(d$weight)) ^ 2) + (b_3 * (weight - mean(d$weight)) ^ 3)) %>% 
  ggplot(aes(x = weight, y = height, group = n)) +
  geom_line(size = 0.1) +
  theme_bw() +
  labs(title = "b ~ dnorm(0, 10)")
```

## Fit the model

```{r}

d %<>%
  mutate(weight_s = (weight - mean(weight, na.rm = T)) / sd(weight, na.rm = T),
         weight_s2 = weight_s ^ 2,
         weight_s3 = weight_s ^ 3)

model <- brm(
  formula = height ~ 1 + weight_s + weight_s2 + weight_s3,
  data = d,
  family = gaussian,
  prior = c(prior(normal(178, 100), class = Intercept),
            prior(lognormal(0, 1), class = b, coef = "weight_s"),
            prior(normal(0, 1), class = b, coef = "weight_s2"),
            prior(normal(0, 1), class = b, coef = "weight_s3"),
            prior(uniform(0, 50), class = sigma)),
  iter = 30000, warmup = 29000, chains = 4, cores = 4,
  seed = 4
)

```

```{r}
plot(model)
```


## And plot it

```{r}
weight_seq %<>%
  mutate(weight_s3 = weight_s ^ 3)

mu_summary <- 
  fitted(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

pred_height <- 
  predict(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

ggplot(data = d, mapping = aes(x = weight_s)) +
  geom_point(aes(y = height)) +
  geom_ribbon(data = pred_height,
              aes(ymin = Q2.5, ymax = Q97.5),
              alpha = 0.5) +
  geom_smooth(data = mu_summary,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "red",
              size = 1,
              alpha = 0.5,
              color = "black") +
  theme_bw()
```

# What about a linear model on the entire data?

## Simulate the priors

```{r}
set.seed(2971)
n_lines <- 100

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b = rlnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b), weight = seq(4, 63, length.out = 20)) %>% 
  mutate(height = a + b * (weight - mean(d$weight))) %>% 
  ggplot(aes(x = weight, y = height)) +
  geom_point(data = d) +
  geom_line(size = 0.1, aes(group = n)) +
  theme_bw() +
  labs(title = "b ~ dnorm(0, 10)")
```

## Fit the model

```{r}

d %<>%
  mutate(weight_s = (weight - mean(weight, na.rm = T)) / sd(weight, na.rm = T))

model <- brm(
  formula = height ~ 1 + weight_s,
  data = d,
  family = gaussian,
  prior = c(prior(normal(178, 100), class = Intercept),
            prior(lognormal(0, 1), class = b, coef = "weight_s"),
            prior(uniform(0, 50), class = sigma)),
  iter = 40000, warmup = 39000, chains = 4, cores = 4,
  seed = 4
)

```

```{r}
plot(model)
```


## And plot it

```{r}


mu_summary <- 
  fitted(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

pred_height <- 
  predict(model, newdata = weight_seq) %>% 
  data.frame() %>% 
  bind_cols(weight_seq)

ggplot(data = d, mapping = aes(x = weight_s)) +
  geom_point(aes(y = height)) +
  geom_ribbon(data = pred_height,
              aes(ymin = Q2.5, ymax = Q97.5),
              alpha = 0.5) +
  geom_smooth(data = mu_summary,
              aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
              stat = "identity",
              fill = "red",
              size = 1,
              alpha = 0.5,
              color = "black") +
  theme_bw()
```

# Splines

```{r}
data(cherry_blossoms)
d <- cherry_blossoms
```

## Visualize the data

```{r, fig.height = 2, fig.width = 6}
ggplot(data = d,
       mapping = aes(x = year, y = temp)) +
  geom_line() +
  geom_point()
```


```{r}

```
















