---
title: "week 2 HW"
author: "Juan Carlos Villaseñor-Derbez"
date: "12/11/2020"
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(rethinking)
  library(brms)
  library(tidyverse)
  library(magrittr)
})
```

# Predict missing heights for given weights

The given weight data are 45, 40, 65, and 31.

```{r}
# Load data
data(Howell1)
# Keep adults only, and center/scale data
d2 <- Howell1 %>% 
  filter(age > 18) %>% 
  mutate(weight_s = (weight - mean(weight, na.rm = T)) / sd(weight, na.rm = T))

# Define missing data, will be useful for plotting and predicting
weights_missing <- c(45, 40, 65, 31)

ggplot(data = d2, aes(x = weight_s, y = height)) +
  geom_point() +
  geom_vline(xintercept = (weights_missing - mean(d2$weight)) / sd(d2$weight), linetype = "dashed") +
  theme_bw() +
  labs(title = "Height (cm) and weight (kg) for !Kung adults",
       subtitle = "Data from rethinking::Howell1",
       caption = "Dashed lines represent location of weight for which height is missing",
       x = "Scaled weight",
       y = "Height (cm)")
```

Define the model to fit

$$
h_i ~ (\mu_i, \sigma) \\
\mu_i = \alpha + \beta \times \text{weight_s}_i \\
\alpha \sim \text{Normal}(195, 50) \\
\beta \sim \text{Log-Normal}(0, 1) \\
\sigma \sim \text{Cauchy}(0, 1)
$$
Simulate from the prior

```{r}
n_lines <- 100
weight_ranges_s <- range((d2$weight - mean(d2$weight)) / sd(d2$weight))

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 195, sd = 50),
       b = rlnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b), weight_s = weight_ranges_s) %>% 
  mutate(height = a + b * weight_s) %>% 
  ggplot(aes(x = weight_s, y = height)) +
  geom_line(aes(group = n),
            size = 0.5,
            alpha = 0.5) +
  geom_point(data = d2)
```

Now actually fit the model

```{r}
adult_model <- brm(formula = height ~ 1 + weight_s,
                   data = d2,
                   family = gaussian,
                   prior = c(prior(normal(195, 50), class = Intercept),
                             prior(lognormal(0, 1), class = b),
                             prior(cauchy(0, 1), class = sigma)),
                   chains = 5,
                   cores = 5,
                   seed = 4)
```

```{r}
plot(adult_model)
```

```{r}

new_data <- tibble(weight = weights_missing) %>% 
  mutate(weight_s = (weight - mean(d2$weight)) / sd(d2$weight))

predict(adult_model, newdata = new_data, probs = c(0.055, 0.945)) %>% 
  data.frame() %>% 
  bind_cols(new_data) %>% 
  select(weight, weight_s, everything(), -Est.Error)

```


# Linear-log model between height and log(weight)


Define some data

```{r}
d <- Howell1 %>% 
  mutate(log_weight = log(weight))

ggplot(data = d, mapping = aes(x = log_weight, y = height)) +
  geom_point() 
```
 Define model
 
$$
h_i \sim \text{Normal}(\mu_i, \sigma) \\
\mu_i = \alpha + \beta \times log(\text{weight}) \\
\alpha = \text{Normal}(100, 50) \\
\beta = \text{Log-Normal}(1, 1) \\
\sigma = \text{Cauchy}(0, 1)
$$

Simulate form the priors

```{r}
n_lines <- 100
log_weight_ranges <- range(d$log_weight)

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 100, sd = 50),
       b = rlnorm(n = n_lines, mean = 1, sd = 1)) %>% 
  expand(nesting(n, a, b), log_weight = log_weight_ranges) %>% 
  mutate(height = a + b * log_weight) %>% 
  ggplot(aes(x = log_weight, y = height)) +
  geom_line(aes(group = n),
            size = 0.5,
            alpha = 0.5) +
  geom_point(data = d)
```


Fit the model

```{r}
all_kung <- 
  brm(formula = height ~ 1 +  log_weight,
      data = d,
      family = gaussian,
      prior = c(prior(normal(100, 50), class = Intercept),
                prior(lognormal(1, 1), class = b),
                prior(cauchy(0, 1), class = sigma)),
      chains = 5,
      cores = 5,
      seed = 4)
```

```{r}
plot(all_kung)
```

```{r, fig.width = 3, fig.height = 3}

mu <- fitted(all_kung, probs = c(0.055, 0.945)) %>% 
  data.frame() %>% 
  mutate(log_weight = d$log_weight,
         weight = d$weight)

height <- predict(all_kung, probs = c(0.055, 0.945)) %>% 
  data.frame() %>% 
  mutate(log_weight = d$log_weight,
         weight = d$weight)

ggplot(mapping = aes(x = weight)) +
  geom_ribbon(data = height, aes(ymin = Q5.5, ymax = Q94.5),
              alpha = 0.5) +
  geom_smooth(data = mu,
              aes(y = Estimate, ymin = Q5.5, ymax = Q94.5),
              stat = "identity",
              color = "black",
              fill = "red",
              alpha = 0.5) +
  geom_point(data = d, aes(y = height),
             size = 1) +
  theme_bw()

```

# Plot prior predictive distribution

For a Quadratic model

$$
\begin{align}
h_i &\sim Normal(\mu_i, \sigma) \\
\mu_i &= a + \beta_1weight\_s_i + \beta_2weight\_s_i^2\\
\alpha &\sim Normal(178, 20)\\
\beta_1 &\sim Log-Normal(0, 1) \\
\beta_2 &\sim Normal(0, 1) \\
\sigma &\sim Uniform(0, 50)
\end{align}
$$

Simulating from the priors

```{r}
set.seed(2971)
n_lines <- 30

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b_1 = rlnorm(n = n_lines, mean = 0, sd = 1),
       b_2 = rnorm(n = n_lines, mean = 0, sd = 1)) %>% 
  expand(nesting(n, a, b_1, b_2), weight = seq(from = min(d$weight), to = max(d$weight), length.out = 30)) %>% 
  mutate(weight_s = (weight - mean(d$weight)) / sd(d$weight),
         weight_s2 = weight_s ^ 2,
         height = a + b_1 * weight_s + b_2 *weight_s2) %>% 
  ggplot(aes(x = weight_s, y = height)) +
  geom_line(size = 0.1, aes(group = n)) +
  theme_bw() +
  labs(title = "b1 ~ dlnorm(0, 1); b2 ~ dnorm(0, 1)") +
  ylim(55, 270)
```


Try tighter / different priors

```{r}
set.seed(2971)
n_lines <- 30

tibble(n = 1:n_lines,
       a = rnorm(n = n_lines, mean = 178, sd = 20),
       b_1 = rlnorm(n = n_lines, mean = 2, sd = 1),
       b_2 = rlnorm(n = n_lines, mean = 1, sd = 1)) %>% 
  expand(nesting(n, a, b_1, b_2), weight = seq(from = min(d$weight), to = max(d$weight), length.out = 30)) %>% 
  mutate(weight_s = (weight - mean(d$weight)) / sd(d$weight),
         weight_s2 = weight_s ^ 2,
         height = a + b_1 * weight_s - b_2 *weight_s2) %>% 
  ggplot(aes(x = weight_s, y = height)) +
  geom_line(size = 0.1, aes(group = n)) +
  theme_bw() +
  labs(title = "b1 ~ dlnorm(0, 1)") +
  ylim(55, 270)
```











